### 2019-06-12 doc/chapters/challenge_stainless.tex
Corrected english

We can either use the sbt plugin or a JAR file to check the code with Stainless.

@@

Having a fixed version configured in the sbt build file, every developer has the same Stainless features available.
This should prevent incompatibility with new or deprecated features when we use different plugins.

@@

However, there are some drawbacks currently.
For example the sbt plugin does not always report errors.
We will read more about that shortly.

@@

The bigger problem is, due to the different Scala and sbt versions, the error message, after trying to go in a sbt shell.
\begin{verbatim}
  [warn] There may be incompatibilities among your library dependencies; run 'evicted'
         to see detailed eviction warnings.
  [error] java.lang.NoClassDefFoundError: sbt/SourcePosition
  ...
  Project loading failed: (r)etry, (q)uit, (l)ast, or (i)gnore?
\end{verbatim}

Downgrading Bitcoin-S sbt version to 0.13.17 fixes the error but then it can not load some libraries only compiled for newer versions.
So this would take too much time to fix and changes the Bitcoin-S code inadvertent.

The next approach is to use stainless cli instead of sbt.
Running stainless on all source files does not work, because the dependencies are missing.
The parameter \emph{-classpath} can resolve it but the value of this parameter must be the paths to all the dependencies separated by a ':'.
Finally, \emph{core} depends on \emph{secp256k1jni}, another package of Bitcoin-S written in Java.
So this needs to be in the source files to.

@@

\emph{.ivy2} is the dependency cache of sbt.
The \emph{tr} replaces the first char with the second so '\\n' with either ':' or ' '.

@@

So we can not know how many errors will face us.
Let's go another way, because the errors may take too much time and it might lead to a next error.
We extract the code needed to verify a transaction mainly the class Transaction and ScriptInterpreter with many other classes their depending on.

After this extraction Stainless was successfully integrated with both, sbt and JAR.

Running \code{sbt compile} in the project with Stainless ended without error.
But it also ended with no output.
So we can not able to change the code so Stainless would accept it since we do not know what to change.

>>>>

We can either use the sbt plugin or a JAR file to check code with Stainless.

@@

Having a static version configured in the sbt build file, every developer has the same Stainless features available.
This should prevent incompatibility with new or deprecated features when we use different plugins.

@@

However, currently there are some drawbacks.
For example the sbt plugin does not always report errors.
We will see more about that shortly.

@@

The bigger problem is, due to the different Scala and sbt versions, the following error after trying to go in a sbt shell:
\begin{verbatim}
  [warn] There may be incompatibilities among your library dependencies; run 'evicted'
         to see detailed eviction warnings.
  [error] java.lang.NoClassDefFoundError: sbt/SourcePosition
  ...
  Project loading failed: (r)etry, (q)uit, (l)ast, or (i)gnore?
\end{verbatim}

Downgrading Bitcoin-S sbt version to 0.13.17 fixes the error but then it can not load some libraries only compiled for newer versions.
So this would take too much time to fix and changes the Bitcoin-S code inadvertently.

The next approach is to use the stainless cli instead of sbt.
Running stainless on all source files does not work, because dependencies are missing.
The parameter \emph{-classpath} can resolve it but the value of this parameter must be the paths to all the dependencies separated by a ':'.
Finally, \emph{core} depends on \emph{secp256k1jni}, another package of Bitcoin-S written in Java.
So this needs to be in the source files to.

@@

\emph{.ivy2} is the dependency cache of sbt.
The \emph{tr} replaces the first char with the second so a newline with either ':' or ' '.

@@

So we can not know how many errors will face us.
Let's go another way, because the errors may take too much time and it might lead to a next error.
We extract the code needed to verify a transaction mainly the class Transaction and ScriptInterpreter with many other classes they're depending on.

After this extraction Stainless was successfully integrated with both sbt and JAR.

Running \code{sbt compile} in the project with Stainless ended without error.
But it also ended with no output.
So we are not able to change the code so Stainless would accept it since we do not know what to change.

### 2019-06-12 doc/chapters/conclusion.tex
Corrected english

\section{Future Work}

>>>>

As we saw, it is really hard to verify software not written with verification in mind.
It needs a lot of changes in the software because verification is mathematical and the current software is written mostly in object-oriented style.
Software written in the functional paradigm would be much easier to reason about.

Thus, either Stainless must find ways to translate more of built-in object-oriented patterns of Scala to their verification tool or developers must invest more in functional programming.

The code we verified in the previous chapter is not the original code from Bitcoin-S.
So we can not guarantee the correctness of the addition of Satoshis with zero in Bitcoin-S.
Not all changes we made were as trivial as the replacement of objects with case objects.
For these non-trivial changes, as seen in section \ref{sec:bound_check}, we cannot say whether they are equivalent to the original implementation or not.

### 2019-06-12 doc/chapters/introduction.tex
Corrected english

With \textit{require} we define a precondition of a function, which we want to verify.
\textit{require} is placed at the beginning of the function body.
\textit{require} specifies, if the parameter of the function hold a certain condition.

With \textit{ensuring}, we define the postcondition of a function.
\textit{ensuring} is placed at the end of the function after the body.
\textit{ensuring} specifies, if the return value or the result of the function hold a certain condition.

@@

The following example demonstrates a simple formal specification for the function calculating factorial.
This is a modified example from the section \href{https://epfl-lara.github.io/stainless/intro.html}{Introduction} of the Stainless documentation \cite{Stainless:documentation}, so Stainless reports an error verifying it.
\begin{lstlisting}[language=Scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}

It recursively calculates factorial of an integer number.
The input of the function is constrained with \textit{require} to a non-negative value.
The result should also be non-negative.
Thus, Stainless will verify whether the result of the factorial calculation is non-negative for all non-negative inputs.

@@

Stainless supports only a subset of Scala.
They call is Pure Scala.
You can find the specification of Pure Scala on the Stainless documentation website \cite{Stainless:documentation} in the section \href{https://epfl-lara.github.io/stainless/purescala.html}{Pure Scala}.

In addition, Stainless has its own library with annotations, reimplementaion of some core data types, collections and input-output functions and more.
Some of them are described on the website in the section \href{https://epfl-lara.github.io/stainless/library.html}{Stainless Library}.
There are more details about library in the source code of Stainless on GitHub \cite{Stainless:github} in the folder \href{https://github.com/epfl-lara/stainless/tree/master/frontends/library/stainless}{\textit{frontends/library/stainless}}.

@@

Bitcoin-S is a large project found on GitHub implementing many functionalities.
In this work we look at two of them, namely the \nameref{property_1} and the \nameref{property_2} described shortly.
Thus, we go through the relevant code parts to verify those properties, look at some challenges with Stainless and see why we should write software with verification in mind.
We will rewrite the code to Pure Scala, because Bitcoin-S is not written in it.

@@

In Bitcoin-S there is a class Satoshis representing an amount of bitcoins.
We look at the verification of the addition of Satoshis with zero Satoshis.
This operation should result in the same amount of Satoshis.
Let's call it the \nameref{property_2}.

>>>>

With \textit{require} we define a precondition of a function, which we want to verify.
\textit{require} is placed at the beginning of the function body.
\textit{require} specifies, if the parameter of the function holds against a certain condition.

With \textit{ensuring}, we define the postcondition of a function.
\textit{ensuring} is placed at the end of the function after the body.
\textit{ensuring} specifies, if the return value or the result of the function satisfies a certain condition.

@@

The following example demonstrates a simple formal specification for the function calculating a factorial.
This is a modified example from the section \href{https://epfl-lara.github.io/stainless/intro.html}{Introduction} of the Stainless documentation \cite{Stainless:documentation}, so Stainless reports an error verifying it.
\begin{lstlisting}[language=Scala]
  def factorial(n: Int): Int = {
      require(n >= 0)
      if (n == 0) {
        1
      } else {
        n * factorial(n - 1)
      }
  } ensuring(res => res >= 0)
\end{lstlisting}

It recursively calculates the factorial of an integer number.
The input of the function is constrained with \textit{require} to a non-negative value.
The result should also be non-negative.
Thus, Stainless will verify whether the result of the factorial calculation is non-negative for all non-negative inputs.

@@

Stainless supports only a subset of Scala.
They call this Pure Scala.
You can find the specification of Pure Scala on the Stainless documentation website \cite{Stainless:documentation} in the section \href{https://epfl-lara.github.io/stainless/purescala.html}{Pure Scala}.

In addition, Stainless has its own library with annotations, reimplementation of some core data types, collections and input-output functions and more.
Some of them are described on the website in the section \href{https://epfl-lara.github.io/stainless/library.html}{Stainless Library}.
There are more details about the library in the source code of Stainless on GitHub \cite{Stainless:github} in the folder \href{https://github.com/epfl-lara/stainless/tree/master/frontends/library/stainless}{\textit{frontends/library/stainless}}.

@@

Bitcoin-S is a large project found on GitHub implementing many functionalities.
In this work we look at two of them, namely the \nameref{property_1} and the \nameref{property_2} described shortly.
Thus, we go through the relevant code parts to verify those properties, look at some challenges with Stainless and see why we should write software with verification in mind.
In order to be able to verify Bitcoin-S, we have to rewrite parts of it in the subset of Scala supported by Stainless.

@@

In Bitcoin-S there is a class \texttt{Satoshis} representing an amount of bitcoins.
We look at the verification of the addition of Satoshis with zero Satoshis.
This operation should result in the same amount of Satoshis.
Let's call it the \nameref{property_2}.

### 2019-06-12 doc/chapters/towards_verifying_addition.tex
changed title
Corrected english

After realizing, that it would consume too much time to rewrite the Bitcoin-S code and even the extracted part with checkTransaction, the smallest unit in Bitcoin-S-Core that is worthwhile to verify was extracted.
This could be the addition of two CurrencyUnits.
To make it even easier, the addition of CurrencyUnits with zero.
CurrencyUnits is an abstract class in Bitcoin-S, representing currencies like Satoshis.

@@

Here, we can just change the objects from object to case object.
Stainless recommendation is to use objects for modules and case objects as algebraic data type.

This is, due to the internal design of Scala.
It's possible to reason about case object but not about object.
This needs a fundamental knowledge of Scala and some functional paradigms that should not be part of this thesis.
The \href{https://github.com/epfl-lara/stainless/issues/520}{issue \mypound520} on Stainless GitHub gives some thoughts if you want to know more.

@@

This should be easy to rewrite by using generics instead of an abstract type, right?
Unfortunately not.
The problem is, CurrencyUnit uses one of its implementing class Satoshis.

@@

As described before, Stainless supports only a subset of Scala.
The BigInt from the Stainless library is a bit restricted.
One such restriction is, that BigInt does not support dynamic BigInt construction.
Thus, the constructor parameter of BitInt must be a literal arguments.

@@

This would be really hard to refactor, because Bitcoin-S tries to be as much dynamic as possible so it can be used with other cryptocurrencies than bitcoins.
Maybe it would even be impossible, because they need to parse dynamic values from the bitcoin network.

Luckily, we can use toBigInt on underlying directly instead of going over toLong.
So, instead of converting the underlying to Long and and back to BigInt we convert underlying directly to BigInt.

@@

This is a missing feature in Stainless.
It does not support upper type bound on the class itself.
To track this, \href{https://github.com/epfl-lara/stainless/issues/519}{issue \mypound519} was created on GitHub during this work.
\begin{lstlisting}[language=scala]
  sealed abstract class Number[T <: Number[T]] extends BasicArithmetic[T]  
\end{lstlisting}

For the sake of convenience we make this a concrete type by replacing T with Int64.
Int64,  because Satoshis uses only Int64.
There are other number typs like UInt16 but for our property we don't need them.

@@

This is easy.
We just replace all occurrence of A with BigInt, since A is not overrode in an implementing class.
This is not exact the same code, because an implementing class can not override A anymore but that's fine for our verification.

@@

This is a bound check.
It checks, if the result of the addition is in range of the specified type, which is now the hard coded Int64.

@@

Seems like Stainless does not support the second string parameter of require or at least it throws a warning about it.
We can safely remove the string parameters from the requires, since they only serve as error message.

@@

Looks like it can not compare a BigInt with a Long value.
We can easily convert this Long value to a BigInt with string literal parameter by passing this numbers as strings to the BigInt constructor.

@@

This show, that there is an invalid specification in checkResult and Stainless prints a counterexample for it.

@@

\section{Writing Specification for Verification}

>>>>

After realizing that it would consume too much time to rewrite the Bitcoin-S code and even the extracted part with checkTransaction, the smallest unit in Bitcoin-S-Core that is worthwhile to verify was extracted.
This could be the addition of two CurrencyUnits.
To make it even easier, the addition of CurrencyUnits with zero.
CurrencyUnits is an abstract class in Bitcoin-S, representing currencies like Satoshis.

@@

Here, we can just change the objects from object to case object.
Stainless recommendation is to use objects for modules and case objects as algebraic data types.

This is due to the internal design of Scala.
It's possible to reason about case object but not about object.
This needs a fundamental knowledge of Scala and some functional paradigms that should not be part of this thesis.
The \href{https://github.com/epfl-lara/stainless/issues/520}{issue \mypound520} on Stainless GitHub gives some thoughts if you want to know more.

@@

This should be easy to rewrite by using generics instead of an abstract type, right?
Unfortunately not.
The problem is, CurrencyUnit uses one of its implementing classes: Satoshis.

@@

As described before, Stainless supports only a subset of Scala.
The BigInt from the Stainless library is a bit restricted.
One such restriction is, that BigInt does not support dynamic BigInt construction.
Thus, the constructor parameter of BitInt must be a literal argument.

@@

This would be really hard to refactor, because Bitcoin-S tries to be as dynamic as possible so it can be used with cryptocurrencies other than bitcoins.
Maybe it could be impossible, because they need to parse dynamic values from the bitcoin network.

Luckily, we can use toBigInt on the field \texttt{underlying} directly instead of toLong.
So, instead of converting the underlying to Long and back to BigInt we convert underlying directly to BigInt.

@@

This is a missing feature in Stainless.
It does not support upper type boundaries on the class itself.
To track this, \href{https://github.com/epfl-lara/stainless/issues/519}{issue \mypound519} was created on GitHub during this work.
\begin{lstlisting}[language=scala]
  sealed abstract class Number[T <: Number[T]] extends BasicArithmetic[T]  
\end{lstlisting}

Despite this, in order to be able to continue, we make this a concrete type by replacing T with Int64.
Int64,  because Satoshis uses only Int64.
There are other number types like UInt16 but for our property we don't need them.

@@

This is easy.
We just replace all occurrence of A with BigInt, since A is not overwritten in an implementing class.
This is not the exact same code, because an implementing class can not override A anymore but that's fine for our verification.

@@

This is a bounds check.
It checks if the result of the addition is in range of the specified type, which is now the hard coded Int64.

@@

Seems like Stainless does not support the second string parameter of require or at least it throws a warning about it.
We can safely remove the string parameters from the requires, since they only serve as error messages.

@@

Looks like it can not compare a BigInt with a Long value.
We can easily convert this Long value to a BigInt with a string literal parameter by passing these numbers as strings to the BigInt constructor.

@@

This shows that there is an invalid specification in checkResult and Stainless prints a counterexample for it.

@@

\section{Writing Specification for the Property}

### 2019-06-12 doc/chapters/trying_verify_checktx.tex
Corrected english

This chapter describes the part of Bitcoin-S needed to verify the \nameref{property_1} described before.
We are going to create a transaction and show you the relevant parts of the method checkTransaction, where transactions are checked against some properties.
Then, we will see the bug in Bitcoin-S found during this work and its fix.
In the end we see why the \nameref{property_1} needed to be changed to the \nameref{property_2}.

@@

Some code in this section is copied or adapted from Bitcoin-S-Core transaction builder example.\cite{BitcoinSCore:txbuilderexample}
Bitcoin-S-Core has a bitcoin transaction builder class with the following constructor:
\begin{lstlisting}[language=scala]
  BitcoinTxBuilder(
    destinations: Seq[TransactionOutput], // where to send the money
    utxos: BitcoinTxBuilder.UTXOMap,      // unspent transaction outputs
    feeRate: FeeUnit,                     // fee rate per byte
    changeSPK: ScriptPubKey,              // where to send the change
    network: BitcoinNetwork               // bitcoin network information
  ): Future[BitcoinTxBuilder]
\end{lstlisting}

The return type Future does not make sense here, since the implementing constructor calls either Future.successful or Future.fromTry which returns an already resolved Future.
This might be for future purposes.

@@

First, we need a some money.
Thus, we create a fake transaction with one single output.
This transaction can be parsed from the bitcoin network, but we create one manually in order to see this process.
\begin{lstlisting}[language=scala]
  val privKey = ECPrivateKey.freshPrivateKey
  val creditingSPK = P2PKHScriptPubKey(pubKey = privKey.publicKey)

@@

Finally, we define the fee rate in satoshis per one byte transaction size as well as some bitcoin network parameters.
The bitcoin network parameters are not important, so we use some static values for testing purpose.
\begin{lstlisting}[language=scala]
  val feeRate = SatoshisPerByte(Satoshis.one)

@@

We can pass a transaction and it returns a Boolean whether the transaction is valid or not.
So for example when we pass the transaction built before the returned value would be true, because it's a valid transaction.
It might not be accepted by the bitcoin network but for a transaction on its own its valid.
We can not check context with it, because we can only pass one transaction.

@@

Here the relevant code of checkTransaction again:
\begin{lstlisting}[language=scala]
  val prevOutputTxIds = transaction.inputs.map(_.previousOutput.txId)
  val noDuplicateInputs = prevOutputTxIds.distinct.size == prevOutputTxIds.size
\end{lstlisting}

What happens, if we have two TransactionOutPoints (previousOutputs) with a different index but referencing the same Transaction ID (txId)?

@@

This was fixed in \href{https://github.com/bitcoin-s/bitcoin-s/pull/435}{pull request \mypound435} on GitHub at April 23, 2019, through this work along a unit test to prevent this bug from appearing again in the future.

@@

Trying to integrate Stainless in Bitcoin-S caused a lot of troubles, mainly because of version conflicts.
You can find more about it in chapter \ref{chap:appendix_arb}.

After integrating Stainless in Bitcoin-S, there were many errors.
It takes too much time to fix them all so it should be easier to extract the classes needed for checkTransaction.

>>>>

This chapter describes the part of Bitcoin-S needed to verify the \nameref{property_1} described before.
We are going to create a transaction and show the relevant parts of the method \texttt{checkTransaction}, where transactions are checked against some properties.
Then, we will see the bug in Bitcoin-S found during this work and its fix.
In the end we see why the \nameref{property_1} needed to be changed to the \nameref{property_2}.

@@

Some code in this section is copied or adapted from the Bitcoin-S-Core transaction builder example \cite{BitcoinSCore:txbuilderexample}.
Bitcoin-S-Core has a bitcoin transaction builder class with the following constructor:
\begin{lstlisting}[language=scala]
  BitcoinTxBuilder(
    destinations: Seq[TransactionOutput], // where to send the money
    utxos: BitcoinTxBuilder.UTXOMap,      // unspent transaction outputs
    feeRate: FeeUnit,                     // fee rate per byte
    changeSPK: ScriptPubKey,              // where to send the change
    network: BitcoinNetwork               // bitcoin network information
  ): Future[BitcoinTxBuilder]
\end{lstlisting}

The return type Future does not make sense here, since the implementation calls either Future.successful or Future.fromTry which returns an already resolved Future.
This might be for future purposes.

@@

First, we need some money.
Thus, we create a fake transaction with one single output.
This transaction can be parsed from the bitcoin network, but we create one manually in order to see this process.
\begin{lstlisting}[language=scala]
  val privKey = ECPrivateKey.freshPrivateKey
  val creditingSPK = P2PKHScriptPubKey(pubKey = privKey.publicKey)

@@

Finally, we define the fee rate in satoshis per one byte transaction size as well as some bitcoin network parameters.
The bitcoin network parameters are not important, so we use some static values normally used when testing.
\begin{lstlisting}[language=scala]
  val feeRate = SatoshisPerByte(Satoshis.one)

@@

We can pass a transaction and it returns a Boolean indicating whether the transaction is valid or not.
So for example when we pass the transaction we built before the returned value would be true, because it's a valid transaction.
It might not be accepted by the bitcoin network but for a transaction on its own it's valid.
We can not check context with it, because we can only pass one transaction.

@@

Here is the relevant code of checkTransaction again:
\begin{lstlisting}[language=scala]
  val prevOutputTxIds = transaction.inputs.map(_.previousOutput.txId)
  val noDuplicateInputs = prevOutputTxIds.distinct.size == prevOutputTxIds.size
\end{lstlisting}

What happens if we have two TransactionOutPoints (previousOutputs) with a different index but referencing the same Transaction ID (txId)?

@@

This was fixed in \href{https://github.com/bitcoin-s/bitcoin-s/pull/435}{pull request \mypound435} on GitHub at April 23, 2019, through this work along with a unit test to prevent this bug from appearing again in the future.

@@

Trying to integrate Stainless in Bitcoin-S caused a lot of troubles, mainly because of version conflicts.
For more details see chapter \ref{chap:appendix_arb}.

After integrating Stainless in Bitcoin-S, there were many errors.
It takes too much time to fix them all so it should be easier to extract the classes needed for the \texttt{checkTransaction} function.
